Tidying the UCI HAR data set
========================================================

This script uses data from the UCI Machine Learning Repository :
**Human Activity Recognition Using Smartphones Data Set**
http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

*Associated research*

Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. 
"Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine." International Workshop of Ambient Assisted Living (IWAAL 2012). 
Vitoria-Gasteiz, Spain. Dec 2012

### Step 1 : data merge

The UCI data sets is broken down in several files : predictor features, predicted activities, subject IDs. Each of this 3 sets is further broken down into training (70% of entire dataset) and test set (remaining 30%). 

Let's start by loading feature labels from provided file :
```{r}
features <- read.table("features.txt", header=FALSE, stringsAsFactors=FALSE)
names(features) <- c("id","name")
```
Then let's load test data sets :
```{r}
xtest <- read.table("test/X_test.txt")
names(xtest) <- features$name
ytest <- read.table("test/y_test.txt")
names(ytest) <- c("activity_id")
stest <- read.table("test/subject_test.txt")
names(stest) <- c("subject_id")
```
Merge test sets side by side :
```{r}
test <- cbind(stest, xtest, ytest)
```
Let's check dimensions and characteristics of our test data :
```{r}
str(test)
```

Load training sets :
```{r}
xtrain <- read.table("train/X_train.txt")
names(xtrain) <- features$name
ytrain <- read.table("train/y_train.txt")
names(ytrain) <- c("activity_id")
strain <- read.table("train/subject_train.txt")
names(strain) <- c("subject_id")
```
Merge training sets side by side :
```{r}
train <- cbind(strain, xtrain, ytrain)
```
Let's check dimensions and characteristics of our training data :
```{r}
str(train)
```
Now let's append training and test sets together (on top of each other, by rows) :
```{r}
full <- rbind(train,test)
str(full)
```

### Step 2 - select specific features

Pattern matching on columns names can be done with matchcols function in gdata package.
```{r}
library(gdata)
```
Match features whose name contain string "mean()" or string "std()" using a regular expression (we make sure to use double backslash to escape parenthesis) :
```{r}
meanstdcols <- matchcols(full, with=c("mean\\(\\)|std\\(\\)"))
```
Ok now let's build a lighter dataset with subject, activity and only mean and std features :
```{r}
sams <- full[,c("subject_id", "activity_id", meanstdcols)]
```

### Step 3 - add activity labels

Load activty labels from provided file :
```{r}
activity <- read.table("activity_labels.txt", stringsAsFactors=FALSE)
names(activity) <- c("activity_id","activity_label")
```
Include explicit activity labels in dataset by merging on activity id :
```{r}
samsExplicit <- merge(sams, activity, by="activity_id")
```

### Step 4 - makes feature names more explicit

After checking on the forums there is consensus about a possible typo in the question (see https://class.coursera.org/getdata-002/forum/thread?thread_id=28#post-461). I could have blindly applied gsub to all names. I decided to be a bit more selective and chose to loop only over feature names to make them more explicit.
```{r}
originalNames <- names(samsExplicit)
newNames <- vector('character')
dontTouch <- c("activity_id","subject_id","activity_label")

for( i in 1:length(originalNames) ){
  name <- originalNames[i]
  # tidy up only if current column name is a feature column name
  # use naive regular expressions to perform substitutions
  if( ! name %in% dontTouch ) {
      a <- gsub("^t", "Time Domain ", name )
      a <- gsub("^f", "Frequency Domain ", a)
      a <- gsub("std\\(\\)", "Standard Deviation ", a)
      a <- gsub("mean\\(\\)", "Mean ", a)
      a <- gsub("-", "", a)
      a <- gsub("Acc", "Linear Acceleration ", a)
      a <- gsub("Jerk", "Jerk ", a)
      a <- gsub("Gyro", "Angular Velocity ", a)
      a <- gsub("Mag","Magnitude ", a)
      a <- gsub("Body", "Body ", a)
      a <- gsub("Gravity", "Gravity ", a)
      newNames <- c(newNames,a)
  }
}
```
Now we can replace our old column names with the new ones.
```{r}
colnames(samsExplicit) <- c("activity_id","subject_id",newNames,"activity_label")
```
Quick check of what our column names look like now :
```{r}
names(samsExplicit)
```

### Step 5 - compute aggregates and save 

Let's create a new dataset showing means of all feature columns grouped by subject_id and activity_id. Since there are 30 individuals and 6 activities, we should end up with 180 rows in the result data frame. We specify feature colums with robust named exclusion rather than specific indices inclusion.
```{r}
samsAggregate <- aggregate( samsExplicit[,-which(names(samsExplicit) %in% dontTouch)], 
                            by=list(samsExplicit$activity_id, samsExplicit$subject_id),
                            FUN=mean)
```
Checking row count:
```{r}
nrow(samsAggregate)
```
Checking our column names are preserved:
```{r}
names(samsAggregate)
```
Bummer, the aggregate function renamed our grouped by cols, let's fix this.
```{r}
names(samsAggregate)[names(samsAggregate)=="Group.1"] <- "activity_id"
names(samsAggregate)[names(samsAggregate)=="Group.2"] <- "subject_id"
head(names(samsAggregate))
```
All good, we can now write our tidy aggregates to text file and upload this dataset for peer assessment.
```{r}
write.table(samsAggregate,file="tidyAggregates.txt",sep=" ",col.names=TRUE,row.names=FALSE)
```
