Impact of severe weather on health and the US economy
========================================================

This is a report demonstrating the impact of severe weather events on population health and the economy at large in the USA. The *Storm Data* we explore is provided by the National Oceanic and Atmospheric Administration (NOAA) as a series of measurements taken between the year 1950 and November 2011.

## Data Processing 

```{r cache=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",destfile="StormData.csv.bz2",method="curl")
data <- read.csv(bzfile("StormData.csv.bz2"),stringsAsFactors=FALSE)
```

Let's give a look at the structure of our dataset :
```{r cache=TRUE}
str(data)
head(data)
```

We will focus on the following variables :
- EVTYPE, weather event type is a string that we should convert as a factor. Three different periods can be identified for which only specific events were reported :
  - 1950-1955 : Tornado
  - 1955-1996 : Tornado, Thunderstorm wind, Hail
  - 1996-2011 : all 48 event types as specified in Directive 10-1605
- FATALITIES, INJURIES : Impact on population in terms of injuries and casualties 
- PROPDMG,PROPDMGEXP : Impact on the economy in terms of property damage
- CROPDMG, CROPDMGEXP : Impact on the economy in terms of crop damage

## Results

We will try to use the dataset to answer 2 questions.

### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

### Across the United States, which types of events have the greatest economic consequences?


