Impact of weather events on health and the US economy
========================================================

This is a report demonstrating the impact of severe weather events on population health and the economy at large in the USA. The *Storm Data* we explore is provided by the National Oceanic and Atmospheric Administration (NOAA) as a series of measurements taken between the year 1950 and November 2011.

# Data Processing 

```{r cache=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",destfile="StormData.csv.bz2",method="curl")
data <- read.csv(bzfile("StormData.csv.bz2"),stringsAsFactors=FALSE)
```

Let's give a look at the structure of our dataset :
```{r cache=TRUE}
str(data)
head(data)
```

We will focus on the following variables :
- EVTYPE, weather event type is a string that we should convert as a factor. Three different periods can be identified for which only specific events were reported :
  - 1950-1955 : Tornado
  - 1955-1996 : Tornado, Thunderstorm wind, Hail
  - 1996-2011 : all 48 event types as specified in Directive 10-1605
- FATALITIES, INJURIES : Impact on population in terms of injuries and casualties 
- PROPDMG,PROPDMGEXP : Impact on the economy in terms of property damage
- CROPDMG, CROPDMGEXP : Impact on the economy in terms of crop damage

It seems we have several timezones
I would be
How dirty is the Hour/minute part?

```{r}
unique(data$BGN_TIME)
```
 - odd use of military time jointly with AM/PM suffix
 - errors such as "1990", "2090", "2400", "9999", "000", "0572", "1580", "13O0" 


```{r}
unique(data$BGN_DATE)
```


```{r}
e <-  data.frame()
for( tz in unique(data$TIME_ZONE)) {
  d <- subset(data,data$TIME_ZONE==tz)
  d$BGN_DATETIME <- as.Date(paste(d$BGN_DATE," ",d$BGN_TIME),format="%m/%d/%Y 0:00:00 %H%M",tz=tz)
  e <- rbind(d,e)
}
data$BGN_DATE <- as.Date(data$BGN_DATE,format="%m/%d/%Y 0:00:00")
```

# Results

## Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

## Across the United States, which types of events have the greatest economic consequences?


